# Reproducible Research
Assignment 1
=========================================

```{r setupoptions, echo = T, results = "hide"}
# setting options in the markdown document
library(knitr)
opts_chunk$set(echo = TRUE)
opts_chunk$set(message = F)
opts_chunk$set(warning = F)
```

This report was written to accomplish the first assignment from Coursera's
*Reproducible Research* course.

## 1. Setting options, loading required packages

```{r options, results = "hide"}
library(ggplot2)
library(dplyr)
library(grid)
library(magrittr)
library(lattice)
Sys.setlocale("LC_TIME", "English")
```

## 2. Reading the data

```{r read_data}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = "rr1_data.zip", method="curl")
dateDownloaded <- Sys.time()
dateDownloaded

data <- read.csv(unz("rr1_data.zip", "activity.csv"))
```

## 3. What is mean total number of steps taken per day?

The first task here was to calculate the total number of steps taken per day:

```{r calc_steps}
steps_per_day <- with(data, tapply(steps, date, sum, na.rm = T))
print(steps_per_day)
```

Secondly, the histogram of the total number of steps taken each day was plotted:

```{r hist_steps1}
qplot(steps_per_day, geom = "histogram",
      xlab = "Steps per day") + geom_bar(fill = "brown")
```

Thirdly, mean and median  of the total number of steps taken per day were reported:

```{r mm1}
mm <- data.frame(mean(steps_per_day), median(steps_per_day))
colnames(mm) <- c("mean", "median")
print(mm, row.names = FALSE)
```

## 4. What is the average daily activity pattern?

Here a time-series plot of the 5-minute interval and the average number of steps taken, averaged across all days was obtained:

```{r time-series}
# data preparation
steps_per_interval <- with(data, tapply(steps, interval, mean, na.rm = T))

# plot
plot3 <- qplot(as.numeric(rownames(steps_per_interval)), steps_per_interval,
      colour = "brown")
plot3 <- plot3 + theme(legend.position = "none")
plot3 <- plot3 + geom_line()
plot3 <- plot3 + xlab("Interval") + ylab("Average number of steps")
plot3
```

What is more, an average 5-minute interval containing the maximum number of steps
was derived:

```{r maxstep}
names(steps_per_interval[match(max(steps_per_interval), steps_per_interval)])
```

## 5. Imputing missing values

First of all, the total number of missing values in the given dataset was calculated:

```{r totalNA}
sum(is.na(data))
```

Next, the NAs were filled by taking the means of the 5-minute intervals. The new
dataset was saved as *data2*.

```{r fillNA}
data2 <- data

for (i in 1:length(data$steps)) {
  if(is.na(data$steps[i])){
    data2$steps[i] <- mean(data$steps[data$interval == data$interval[i]], na.rm = T)
  }
}
```

A new histogram of the total number of steps taken each day was created as well:

```{r hist_steps2}
# data preparation
steps_per_day2 <- with(data2, tapply(steps, date, sum, na.rm = T))

# histogram
qplot(steps_per_day, geom = "histogram",
      xlab = "Steps per day") + geom_bar(fill = "brown")
```


New mean and median  of the total number of steps taken per day were also reported:

```{r mm2}
mm2 <- data.frame(mean(steps_per_day2), median(steps_per_day2))
colnames(mm2) <- c("mean", "median")
print(mm2, row.names = FALSE)
```

The obtained values are different from those in the first part. Imputing missings
using means by intervals improved the mean and median values of total daily numbers
of steps.

## 6. Are there differences in activity patterns between weekdays and weekends?

Firstly, a new variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day was created
(it was assumed that a weekend contains Saturday and Sunday):

```{r weekend1}
data2$day <- weekdays(strptime(data2$date, "%Y-%m-%d"))
data2$weekend <- ifelse(data2$day %in% c("Saturday", "Sunday"), "weekend", "weekday")
```

Next, some data transformations were needed in order to obtain a plot by using the
lattice system.

```{r weekend2}
weekday <- 
  data2 %>% subset(weekend == "weekday") %>% with(tapply(steps, interval, mean))

weekend <- 
  data2 %>% subset(weekend == "weekend") %>% with(tapply(steps, interval, mean))

weekday <- data.frame(weekday, as.numeric(rownames(weekday)), "weekday")
weekend <- data.frame(weekend, as.numeric(rownames(weekend)), "weekend")

colnames(weekend) <- c("steps", "interval", "weekend")
colnames(weekday) <- c("steps", "interval", "weekend")

data3 <- rbind(weekday, weekend)
```

Finally, the panel plot containing a time series plot of the 5-minute interval 
(x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis) was created:

```{r weekend3}
xyplot(steps ~ interval | weekend, data = data3, layout = c(1, 2), type = "l",
       col = "brown",
       par.settings = list(strip.background = list(col = "whitesmoke")))
```